Stochastic Blockmodels for Relational Event Data

### 2012-03-23 docs/submission.tex
Put report for Smyth in NIPS format and submit.

\pagestyle{empty}
\documentclass[11pt]{article}
\usepackage{tikz}
\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]

%\usepackage{tikz,fullpage}
%\usetikzlibrary{arrows,petri,topaths}%
%\usepackage{tkz-berge}
\usepackage{verbatim}
\usepackage{url}
\usepackage{graphicx, color}
\usepackage{amsmath}
\usepackage{subfigure}
%\usepackage[small,compact]{titlesec} 
%\usepackage[small,it]{caption}

\textwidth 6.5in
\headheight .1in
\topmargin -.2 in
\textheight 8.7in
\oddsidemargin 0in
\evensidemargin 1in

@@

 \centerline{\Large \bf Stochastic Blockmodels for Relational Event Data} 
 \medskip
\centerline{\bf Christopher DuBois}
 \bigskip

\abstract{
Observing network data as a stream of events provides an opportunity to understand differences in how nodes behave given the past.  Recent methods for continuous-time network data model the rate of dyadic events conditioned on the observed history of events and covariates about the actors.  Methods for static networks, such as stochastic blockmodels, often account for actor-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.  We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.  The method is illustrated using dyadic interaction data -- e.g. email, Twitter direct messages.  We show parameter estimates from the model reveal heterogeneity in the dynamics among groups of individuals.  The fitted models have better predictive accuracy than both baselines and relational event models without the latent structure.  Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.

%Stochastic blockmodels are a tool for exploring the structure of static social networks by learning latent groups of individuals who have similar tendencies in their  group-wise interactions. Continuous-time methods, such as the relational event framework (Butts 2008), model the rate of dyadic events conditioned on the observed history of events and covariates about the actors.  We propose a hierarchical model combining these two approaches where each pair of blocks has a set of parameters guiding the event dynamics for dyads between those blocks.  The  method is illustrated using  dyadic interaction data (e.g. email, Twitter) among hundreds of actors and thousands of events.  We show parameter estimates from the model can reveal heterogeneity in the dynamics among groups of individuals, providing a better fit to the data than a single relational event model for all actors in the network.  Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.

% \begin{itemize}
% \item Blockmodels and clustering algorithms often applied to static networks for exploratory data analysis
% \item continuous time methods allow one to predict future events given the previous history
% \item we combine these two ideas with a hierarchical model such that each pair of blocks has a set of parameters guiding the event dynamics for those dyads
% \item prediction experiments on real datasets
% \item show that the parameter estimates are interpretable
% \end{itemize}

}

@@

Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that assumes latent clusters of nodes share similar patterns of interaction.  The behavior of the model is illustrated with simulated data.  We describe parameter estimation and the learning of the latent cluster assignments via MCMC.  Using data sets of dyadic communication among people, we compare the predictive performance of the fitted models to standard baselines.  Finally we show that the parameter estimates provide an interpretable structure to the event dynamics, sometimes identifying particular \emph{roles} of communication that exist.

@@

One can extend the above to marked point processes where each event contains additional information.  In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.  We assume each dyad $(i,j) \in \mathcal{R}$ has an intensity function $\lambda_{ij}(t)$ that is piecewise constant.  We can then write the likelihood as  
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | z_i,z_j) \}
\label{eqn:llk}
\end{align}
\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{i,j,m}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs. \footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}

@@

% labels
\node at (-.5,2) {$\lambda_{1,2}$};
\node at (-.5,1) {$\lambda_{3,4}$};
\node at (1,-.3) {$t_1$};
\node at (3,-.3) {$t_2$};
\end{tikzpicture}
}
\caption[]{Illustration of event data and the assumptions of the model.  Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.  Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).  Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.\footnotemark }
\label{fig:example}
\end{figure}

@@

For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $\mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z})$ that is a function of the previous event history  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t)\}$.   Though this will be useful, we will constrain the types of dependencies so that an interaction between one dyad does not affect the rate of an entirely separate dyad.  As shown in Figure \ref{fig:example}, we do this by assuming each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.  This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.

We hope to learn about the dynamics between subsets of nodes while allowing for variety among disjoint subsets.   To facilitate this, we assume node $i$ is assigned a latent cluster $z_i$ and the intensity functions via a log linear model
\begin{align*}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) &= \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t) &
z_i &\sim \mbox{CRP}(\alpha) &
\beta_{k,l,p} &\sim \mbox{N}(\mu_p,\sigma_p^2)
\end{align*}
where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the covariate vector $\mathbf{s}(t,i,j|\cdot)$.   Thus, the model of dynamics for the dyad $(i,j)$ has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.  To allow for the blocks to share information by placing a hierarchical prior on each $\boldsymbol{\beta}_{k.l}$.

@@

% \begin{enumerate}
% \item Intercept: $s_{0}(t,i,j) = 1$
% \item Dyad count: $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$
% \item Sender out-degree: $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$
% \item Sender in-degree: $s_{2}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$
% \item Receiver out-degree: $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$
% \item Receiver in-degree: $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$
% \item Reciprocity (AB-BA): $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
% \item Turn-taking (AB-BY): $s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
% \item Turn-usurping (AB-XA): $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
% \item Turn-usurping (AB-XB): $s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
% \item Turn-continuing (AB-AY): $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$
% %\item Number of shared two-paths: $s_{10}(t,i,j) =   \sum_{m:t_m<t} \sum_{r=1}^NI(i_m=i,j_m=r or i_m=r,j_m=j)$
% % \item Shared contacts (since last occurrence): $s_{ij10}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}=r)$
% % \item Triadic closure (since last occurrence): $s_{ij11}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=j)$
% % \item Shared contacts (since last occurrence): $s_{ij12}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{ijm}}=r,j_{v_{ijm}}=r)$
% \end{enumerate}

\begin{table}[t]
\footnotesize
\center
\begin{tabular}{|l|l|}
\hline
Statistic & Formula \\
\hline
\hline
Intercept:& $s_{0}(t,i,j) = 1$\\
Dyad count:& $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
Sender out-degree:& $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
Sender in-degree:& $s_{2}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
Receiver out-degree:& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
Receiver in-degree:& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
Reciprocity (AB-BA):& $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}=i)$\\
Turn-taking (AB-BY): &$s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}}=j,j_{v_{ijm}}\ne i)$\\
Turn-usurping (AB-XA):& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}} \ne j,j_{v_{ijm}}=i)$\\
Turn-usurping (AB-XB): &$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{ijm}} \ne i,j_{v_{ijm}}=j)$\\
Turn-continuing (AB-AY):& $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{ijm}}=i,j_{v_{ijm}}\ne j)$\\
\hline
\end{tabular}
\label{tab:stats}
\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$.}
\end{table}

@@

The above framework generalizes several important special cases. For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.  Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.  As these intensities do not change under this specification, the likelihood simplifies to 
$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$

Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood becomes a conditional logit
$$\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.$$
The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.

\subsection{Related work}

Relational event models \cite{Butts2008} require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant and allow for a nonlinear relationship between statistics and intensity functions.

(TODO: More here.  For example: \cite{Tillman2011} explores the idea of having nonpoisson processes at each edge and solves the master equation for the dynamical system.)

@@

We Gibbs sample the latent class assignments $\mathbf{z}$ from the conditional distribution
\begin{align*}
p(z_i | z_{-i},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto p(z_i | z_{-i},\alpha) p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ i \ \mbox{involved}).% & z_i &= \mbox{new} = \alpha
\end{align*}
Letting $\mathcal{U}_r = \{a: r \in \{i_a,j_a\}, a \in [1,M]\}$ be the set of events where node $r$ is involved, the likelihood term above can be written as
$$p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ r \ \mbox{involved}) = \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}$$

\noindent Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$. 
% \begin{align*}
% p(z_i = k | z_{-i},\alpha) &= n_k & p(z_i = \mbox{new} |  z_{-i},\alpha) &= \alpha
% \end{align*}

@@

\subsection*{Sampling $\boldsymbol{\beta}$ given $\mathbf{z}$ and $\mathcal{A}$ }

For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
\begin{align*}
p(\boldsymbol{\beta}_{k,l} | \mathcal{A}_t, \textbf{z}, \mu, \sigma) &\propto p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) \\  
p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)
\end{align*}
the likelihood term above can be written as
$$p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) = \prod_{a \in \mathcal{V}_{k,l}} \lambda_{i_a,j_a}(t_a|\cdot)
\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}$$
\noindent where $\mathcal{V}_{k,l} = \{a: z_{i_a} \in \{k,l\} \mbox{or} z_{j_a} \in \{k,l\}, a \in [1,M]\}$ is the set of events $a$ where one of the nodes is either in block $k$ or $l$.  \footnote{TODO: If we end up using HMC, mention step size and step length used.}
The above distribution is sampled using slice sampling.\footnote{We may also use HMC for this.}

\subsection*{Split-merge algorithm}

Motivation: simply Gibbs sampling does not seem to mix well.  \footnote{This section is planned future work.}

In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.  For these chains we do not propose split moves when the number of non-empty clusters is already $K$.

@@

\begin{figure}
\center
\includegraphics[width=1.6in]{../figs/synthetic/mat.pdf}
\includegraphics[width=2.25in]{../figs/synthetic/counts.pdf}
\includegraphics[width=2.5in]{../figs/synthetic/params-estimates.pdf}
\caption{Illustration of 1000 simulated events, as described in text. Left: Counts of each dyad. Center: Boxplot of distribution of participation counts across dyads.  The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.  Right: Parameters (in red) and posterior credible intervals (in black).}
\label{fig:syncounts}
\end{figure}

@@

In Figure \ref{fig:syncounts} we illustrate the simulated data.  While the dyad counts suggest a stochastic blockmodel, the center plot shows each block has empirical differences in their dynamics: intensities for reciprocal actions among actors in block 1 are $e^3$ times greater, intensities for turn-taking actions among actors in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.  Fitting the model with $K=2$ converges to the true loglikelihood (see Table \ref{tab:results}) and the posterior credible intervals of the parameters cover the true parameter values.

@@

\small{
\input{../figs/results.tex}
}

\subsection*{Data}

@@

\begin{itemize}
%\item Email: In small organizations there may be groups of people who tend to have a similar style of interaction, and who also share a pattern of interaction with those outside the group or in another group.
\item Eckmann email:
\item Kiel email: Still need to hunt this down.
\item Enron email:
\item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.  We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.  Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.  We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
\item MIT Reality Mining: 
\end{itemize}

\subsection*{Prediction experiment}

\begin{figure}[th]
\caption{Recall plots showing predictive performance on a ranking task.  Left: Eckmann.  Right: Another dataset.  (TODO)}
\label{fig:recall}
\end{figure}

>>>>

\documentclass{article}
\usepackage{nips12submit_e,times}
\usepackage{tikz}
\tikzstyle{vertex}=[auto=left,circle,fill=black!25,minimum size=20pt,inner sep=0pt]
\usepackage{verbatim}
\usepackage{url}
\usepackage{graphicx, color}
\usepackage{amsmath}
\usepackage{subfigure}

\title{Stochastic Blockmodels for Relational Event Data}

\author{
David S.~Hippocampus\thanks{ Use footnote for providing further information
about author (webpage, alternative address)---\emph{not} for acknowledging
funding agencies.} \\
Department of Computer Science\\
Cranberry-Lemon University\\
Pittsburgh, PA 15213 \\
\texttt{hippo@cs.cranberry-lemon.edu} \\
\And
Coauthor \\
Affiliation \\
Address \\
\texttt{email} \\
\AND
Coauthor \\
Affiliation \\
Address \\
}

@@

\maketitle

\begin{abstract}
Observing network data as a stream of events provides an opportunity to understand differences in how nodes behave given the past.  Recent methods for continuous-time network data model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.  Methods for static networks, such as stochastic blockmodels, often account for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.  We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.  The method is illustrated using dyadic interaction data -- e.g. email, Twitter direct messages.  We show parameter estimates from the model reveal heterogeneity in the dynamics among groups of individuals.  The fitted models have better predictive accuracy than both baselines and relational event models without the latent structure.  Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
\end{abstract}

@@

Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that assumes latent clusters of nodes share similar patterns of interaction.  The behavior of the model is illustrated with simulated data.  We describe parameter estimation and the learning of the latent cluster assignments via MCMC.  Using data sets of dyadic communication among people, we compare the predictive performance of the fitted models to standard baselines.  Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, sometimes identifying particular \emph{roles} of communication.

@@

One can extend the above to marked point processes where each event contains additional information.  In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.  We assume each dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t)$ that is piecewise constant.  We can then write the likelihood as  
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{mij}) \lambda_{ij}(t_m | z_i,z_j) \}
\label{eqn:llk}
\end{align}
\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{mij}$ is the time  of the changepoint for $\lambda_{i,j}$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs. \footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}

@@

% labels
\node at (-.5,2) {$\lambda_{1,2}$};
\node at (-.5,1) {$\lambda_{3,4}$};
\node at (1,-.3) {$t_1$};
\node at (3,-.3) {$t_2$};
\end{tikzpicture}
}
\caption[]{Illustration of event data and the assumptions of the model.  Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.  Right: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).  Note $\lambda_{34}$ can only change when events occur that involve either node 3 or node 4.\footnotemark }
\label{fig:example}
\end{figure}

@@

For a given dyad $(i,j)$ and time $t$ we may also have a vector of statistics $\mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z})$ that is a function of the previous event history  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t)\}$.   In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.  As shown in Figure \ref{fig:example}, we incorporate this idea by assuming each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.  This implies that  $\mathcal{R}_{i,j}$ is the set of dyads involving either $i$ or $j$ and that we will only use  statistics $\mathbf{s}(t,i,j)$ that do not change until an event involving either $i$ or $j$ occurs.

We hope to learn about the dynamics between subsets of nodes.   To facilitate this, we assume node $i$ is assigned a latent cluster $z_i$ and the intensity functions via a log linear model
\begin{align*}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) &= \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t) &
z_i &\sim \mbox{CRP}(\alpha) &
\beta_{k,l,p} &\sim \mbox{N}(\mu_p,\sigma_p^2)
\end{align*}
where for each pair of clusters $(z_i,z_j)$ we have a vector of parameters $\boldsymbol{\beta}_{z_i,z_j}$ that corresponds to the vector of statistics $\mathbf{s}(t,i,j|\cdot)$.   Thus, the model of dynamics for the dyad $(i,j)$ has the same parameters as for other dyads occurring between group $z_i$ and $z_j$.  We allow  the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$.

@@

% \begin{enumerate}
% \item Intercept: $s_{0}(t,i,j) = 1$
% \item Dyad count: $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$
% \item Sender out-degree: $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$
% \item Sender in-degree: $s_{2}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$
% \item Receiver out-degree: $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$
% \item Receiver in-degree: $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$
% \item Reciprocity (AB-BA): $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$
% \item Turn-taking (AB-BY): $s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$
% \item Turn-usurping (AB-XA): $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$
% \item Turn-usurping (AB-XB): $s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$
% \item Turn-continuing (AB-AY): $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$
% %\item Number of shared two-paths: $s_{10}(t,i,j) =   \sum_{m:t_m<t} \sum_{r=1}^NI(i_m=i,j_m=r or i_m=r,j_m=j)$
% % \item Shared contacts (since last occurrence): $s_{ij10}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}=r)$
% % \item Triadic closure (since last occurrence): $s_{ij11}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{mij}}=r,j_{v_{mij}}=j)$
% % \item Shared contacts (since last occurrence): $s_{ij12}(t,i,j) = \sum_{r=1}^N I(i_m=i,j_m=j,i_{v_{mij}}=r,j_{v_{mij}}=r)$
% \end{enumerate}

\begin{table}[t]
\footnotesize
\center
\begin{tabular}{|l|l|}
\hline
Statistic & Formula \\
\hline
\hline
Intercept& $s_{0}(t,i,j) = 1$\\
Dyad count& $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
Sender out-degree& $s_{1}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
Sender in-degree& $s_{2}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
Reciprocity (AB-BA)& $s_{5}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
Turn-taking (AB-BY)&$s_{6}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
Turn-continuing (AB-AY)& $s_{9}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
\hline
\end{tabular}
\label{tab:stats}
\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$.}
\end{table}

@@

The above framework generalizes several important special cases. For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.  Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.  As these intensities do not change under this specification, the likelihood simplifies to 
$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$

Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood can be written as
\begin{align}
\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.
\label{eqn:multllk}
\end{align}
The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.

@@

We Gibbs sample the latent class assignments $\mathbf{z}$ from the conditional distribution
\begin{align*}
p(z_i | z_{-i},\mathcal{A},\alpha,\boldsymbol{\beta}) &\propto p(z_i | z_{-i},\alpha) p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ i \ \mbox{involved}) \\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ r \ \mbox{involved}) &= \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
\end{align*}
where $\mathcal{U}_r = \{a: r \in \{i_a,j_a\}, a \in [1,M]\}$ is the set of events where node $r$ is involved.  Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$. 
% \begin{align*}
% p(z_i = k | z_{-i},\alpha) &= n_k & p(z_i = \mbox{new} |  z_{-i},\alpha) &= \alpha
% \end{align*}

@@

\subsection{Sampling $\boldsymbol{\beta}$ given $\mathbf{z}$ and $\mathcal{A}$ }

For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
\begin{align*}
p(\boldsymbol{\beta}_{k,l} | \mathcal{A}_t, \textbf{z}, \mu, \sigma) &\propto p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) \\  
p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) &= \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
\end{align*}
where $\mathcal{V}_{k,l} = \{a: z_{i_a} \in \{k,l\} \ \mbox{or} \ z_{j_a} \in \{k,l\}, a \in [1,M]\}$ is the set of events $a$ where one of the nodes is either in block $k$ or $l$.  \footnote{TODO: If we end up using HMC, mention step size and step length used.}
This distribution is sampled using slice sampling.\footnote{We may also use HMC for this.}

\subsection{Split-merge algorithm}

Motivation: simply Gibbs sampling does not seem to mix well.  \footnote{This section is planned future work. In the experiments we evaluate the performance of the model when restricting the number of clusters $K$.  For these chains we do not propose split moves when the number of non-empty clusters is already $K$.}

@@

\begin{figure}
\center
\includegraphics[width=1.6in]{../figs/synthetic/mat.pdf}
\includegraphics[width=2in]{../figs/synthetic/counts.pdf}
\includegraphics[width=1.8in]{../figs/synthetic/params-estimates.pdf}
\caption{Illustration of 2000 simulated events, as described in text. Left: Counts of each dyad. Center: Boxplot of distribution of participation counts across dyads.  The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.  Right: Parameters (in red) and posterior credible intervals (in black).}
\label{fig:syncounts}
\end{figure}

@@

Though the dyad counts for the synthetic data set suggest a stochastic blockmodel (as seen in Figure  \ref{fig:syncounts}),  the center plot shows each block has empirical differences in their dynamics.  Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.  Fitting the model with $K=2$ converges to the true loglikelihood (see Table \ref{tab:results}) and the posterior credible intervals of the parameters cover the true parameter values.

@@

\subsection{Data}

@@

\begin{itemize}
\item Kiel email: \cite{Ebel2002}
\item Enron email:
\item Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.\footnote{Will be made publicly available.}  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.  We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.  Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.  We use the subset of 487 users who participated in more than one event, using a training set of 2000 events and a test set of 1330 events.
\item MIT Reality Mining: 
\end{itemize}

\subsection{Prediction experiments}
